---
title: "MortaliTeach - build_log"
author: "Debasrija"
date: "2026-05-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)

setwd("/home/ubunt/Downloads/mortaliteach/")
insurance = read.csv("data/insurance.csv")
```

## Step 1 - Data and personas

## Why this project exists

Life and Health insurance pricing is built on this question: given 
everything we know about a person, how likely are they to cost us 
money? The Experience Study is the feedback loop that tells an 
insurer whether their answer to that question was right.

This project builds a tool that teaches that feedback loop from 
first principles, from a single scatter plot to a renegotiated 
reinsurance treaty, using real data and production-grade R engineering.

## What this project teaches

A person who completes this app will understand how insurance 
premiums are derived from observed data, how a life insurance 
pool works, what an Experience Study measures, why count A/E and 
amount A/E can tell different stories, and why reinsurers exist.

## Why this dataset specifically

### What the data is

1,338 real US health insurance records. Age, sex, BMI, smoking 
status, number of children, region, and actual charges billed.

Used to answer: what drives premium differences?

### What I found

- 1,338 rows, 7 columns, no missing values
- Charges: min $1,122, median $9,382, mean $13,270, max $63,770
- Right-skewed, so a small number of very expensive cases pull the mean up
- Age range 18–64
- Approximately equal male-female split, 662 female and 676 male

### What causes the difference in premiums?

One thing that affects premiums significantly is age. Premiums generally go up with age as they are considered higher risk in terms of both life and health.

```{r scatter}
ggplot(data = insurance) + geom_density(aes(x = charges)) + theme_minimal()
ggplot(data = insurance) + geom_point(aes(x = age, y = charges)) + theme_minimal()
```

Interestingly - while the general trend of premiums going up holds, there are 3 bands, which means there is/are some other factor/factors behind the trend, not just age.

#### Smoker split

Based on background reading, I also knew that insurance premiums vary widely depending on whether the customer is a smoker or not, as it affects the mortality measures. So, I decided to start with that.

- There are 274 (20.48%) smokers and 1064 non-smokers. No missing values.
- Mean charges among smokers is $32050.23 and non-smokers is $8434.27, around 3.8x.

Hence, 1,338 real insurance premiums show that smokers pay nearly 3.8 times the premium as compared to a non-smoker.


#### Split by smoking status

```{r scattersmoke}

ggplot(data = insurance) + geom_point(aes(x = age, y = charges, colour = smoker)) +
  scale_color_brewer(palette = "Set1") + theme_minimal()
```
The gap seems partially resolved. The top and bottom bands are composed entirely of smokers and non-smokers, but the middle band is a mixture. There may be other factors which have not been accounted for yet.

#### Split by smoking and obesity status

Another possible split might be BMI (obese/non-obese). Obesity and other comorbidities push premiums high up significantly as well.

```{r scattersmokebmi}

insurance$obese[insurance$bmi >= 30] = "yes"
insurance$obese[insurance$bmi < 30] = "no"

ggplot(data = insurance) + geom_point(aes(x = age, y = charges, colour = smoker, shape = obese), size = 2) +
  scale_color_brewer(palette = "Set1") + theme_minimal()
```

Adding obesity (BMI >= 30) as shape resolves the remaining 
structure. Four groups are now visible:

- Smoker + obese: highest charges, wide spread
- Smoker + non-obese: high charges, below obese smokers  
- Non-smoker + obese: middle band
- Non-smoker + non-obese: bottom band, tight and predictable

Smoking is the dominant signal. Obesity is secondary but 
real — it creates the middle band that was unexplained.

### The four groups

Breaking down by both smoking status and obesity (BMI >= 30):

| Group                    | Mean charges | 
|--------------------------|-------------|
| Smoker + obese           | $41,558     |
| Smoker + non-obese       | $21,363     |
| Non-smoker + obese       | $8,843      |
| Non-smoker + non-obese   | $7,977      |

Two things stand out immediately:

**Smoking dominates obesity.** A smoker with normal BMI 
($21,363) costs more than twice a non-smoker who is obese 
($8,843). The smoking effect overwhelms the BMI effect.

**Obesity matters much more when combined with smoking.** 
The obese/non-obese ratio among smokers is 1.95x ($41,558 
vs $21,363). Among non-smokers it is only 1.11x ($8,843 vs 
$7,977). Obesity and smoking interact - they are not 
simply additive.

These four groups map to risk classes as follows:

- Non-smoker + non-obese  - Preferred Elite
- Non-smoker + obese      - Preferred
- Smoker + non-obese      - Substandard
- Smoker + obese          - Substandard (higher loading)

Smoking alone is disqualifying for Standard or better in real underwriting. Both smoker groups sit in Substandard. The obesity factor makes the combined group more expensive but does not change the classification tier: it changes the premium multiplier within that tier.

I can build a decision tree with this information, grouping subjects into further groups depending on whether this exploration missed any other factors.

So, this mapping may be potentially revised 
when the decision tree is fitted. The tree may suggest 
different natural groupings.

The first iteration of the decision tree is here - 

```{r decisiontree}
library(rpart)
library(rpart.plot)

insurance$smoker <- as.factor(insurance$smoker)
insurance$sex    <- as.factor(insurance$sex)
insurance$region <- as.factor(insurance$region)

tree <- rpart(charges ~ age + bmi + smoker + sex + children + region,
              data = insurance,
              method = "anova")

importance <- as.data.frame(tree$variable.importance)
importance$variable <- rownames(importance)
colnames(importance)[1] <- "importance"

ggplot(importance, aes(x = reorder(variable, importance), 
                       y = importance)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Variable importance in predicting charges",
       x = NULL, y = "Importance") +
  theme_minimal()
```

This shows that smoking status, BMI and age are the three main factors behind the split in charges. I did not add obese as a dichotomous variable since I wanted to see where the split naturally happens given the dataset. (Note for self: charges is a continuous outcome variable, so the method used was ANOVA.)

```{r dt}
printcp(tree)
rpart.plot(tree, type = 4, extra = 101, roundint = FALSE)
```

Within non-smokers, age is more discriminating than obesity at this level of the tree. The younger non-smokers average $5,399, older ones $12,000. Within smokers, the dominant factor becomes obesity, with the premiums being higher than non-smokers all around, non-obese individuals averaging $21,000 and obese averaging $42,000.

Another interesting thing is that the BMI categorisation for smokers does happen at the exact thresholds where obesity is defined. This serves as an extra sanity check as to whether the thresholds used earlier were correct given the dataset.

#### Why sex, region, and children were not used

Sex, region, and children are not predictive of charges once smoking, age, and BMI are known. This is a finding, not a data problem.

In real insurance underwriting, sex does affect pricing (women generally have lower mortality than men). The fact that it does not appear here is likely because this is a
health insurance charges dataset, not a life insurance mortality dataset. 


The next thing to do now is to pick one representative persona for each of these groups. I decided to pick the individuals closest to the mean.


```{r persona}
insurance$leaf <- tree$where

leaf_means <- aggregate(charges ~ leaf, data = insurance,
                        FUN = mean)

personas <- do.call(rbind, lapply(unique(insurance$leaf), function(l) {
  group <- insurance[insurance$leaf == l, ]
  leaf_mean <- leaf_means$charges[leaf_means$leaf == l]
  closest <- which.min(abs(group$charges - leaf_mean))
  group[closest, ]
}))

print(personas)
```

These individuals are real ones picked from the dataset, and reduces the amount of ambiguity in deciding what characteristics should be present in these personas when designed. 

Since I am planning to make an app, I will assign names to these individuals.

Now we give them names: (_wild Vampire Diaries reference appears_...)

- Row 524: **Elena** - 38, female, non-smoker, obese. 
  Preferred Elite despite high BMI because she does not smoke 
  and is under 43. Pays $5,398.

- Row 842: **Damon** - 59, male, non-smoker, normal BMI. 
  Preferred as older age moves him out of Preferred Elite. 
  Pays $12,324.

- Row 161: **Caroline** - 42, female, smoker, normal BMI. 
  Substandard. Smoking alone puts her here regardless of 
  normal BMI. Pays $21,349.

- Row 1091: **Niklaus** - 47, male, smoker, obese. 
  Substandard high loading. Both risk factors present. 
  Pays $41,676.

These four people will appear in every section of the app. 
Their numbers are real. Their names are not. Overall, the message here is that the premium order is driven almost entirely by smoking. This is not a biological truth, but is true for this dataset.

Making a dataframe for these 4 individuals.

```{r personadt}
# VBT 2015 central death rates for the four personas
# Source: SOA Valuation Basic Table 2015
# mx = central death rate (deaths per person per year)

personas_vbt <- data.frame(
  name   = c("Elena", "Damon", "Caroline", "Niklaus"),
  age    = c(38, 59, 42, 47),
  sex    = c("female", "male", "female", "male"),
  smoker = c("no", "no", "yes", "yes"),
  bmi    = c(37.73, 24.70, 26.60, 36.19),
  health_charges = c(5397.617, 12323.936, 21348.706, 41676.081),
  risk_class = c("Preferred Elite", "Preferred", 
                 "Substandard", "Substandard High"),
  stringsAsFactors = FALSE
)
```

ANB is used as the best approximation of age since it generally gives a more precise result. 

From the 2015 VBT lookup table - we can get qx, and assuming uniform force of mortality, mx can be calculated. 

```{r qx}
personas_vbt$qx <- c(
  0.00012, # Elena: F38NS
  0.00378, # Damon: M59NS  
  0.00045, # Caroline: F42S
  0.00323 # Niklaus: M47S
)

personas_vbt$mx = -log(1 - personas_vbt$qx)

print(personas_vbt)
```


Mortality vs charges ratio at similar ages should be consistent, in the same direction at least.
```{r check}
# mortality ratio: smoker vs non-smoker at similar ages
cat("Caroline/Elena mortality ratio:", 
    round(personas_vbt$mx[3] / personas_vbt$mx[1], 2), "\n")

cat("Health charges ratio (smoker/non-smoker):", 
    round(21348 / 5397, 2), "\n")
```

Smoker/non-smoker mortality ratio (3.75x) is consistent 
with the health charges ratio from insurance data (3.96x). 
This confirms the bridge between the two datasets - the 
same variables that predict health costs predict mortality. 

### Why a GLM alongside the decision tree

The decision tree showed us the risk classes visually. It is readable, intuitive, and derives groupings from data without any prior assumptions. But it has some weaknesses that make it unsuitable as a standalone model:

1. Small changes in the data can change the splits entirely.
2. The decision tree cannot adjust for factors like age and BMI between the smokers and non-smokers.
3. Age splits at exactly 43, but the true relationship between age and charges is continuous so it is not explainable fully.
4. No uncertainty estimates like standard errors, or confidence intervals.

The GLM addresses all three. It models the continuous 
relationship between each variable and charges, produces 
standard errors on every coefficient, and is stable across 
small data changes.

### Why Gamma with log link

Charges are strictly positive and right-skewed. A standard 
linear regression (Gaussian GLM) would allow negative 
predicted values, which is impossible for insurance charges. The 
Gamma distribution respects the constraint that charges > 0 
and matches the shape of the observed distribution.

The log link models multiplicative effects. A smoker does 
not pay a fixed dollar amount more than a non-smoker, they pay a multiple of what a non-smoker pays. The log link captures this correctly. It also means coefficients 
are directly interpretable as multipliers when exponentiated:

exp(coefficient) = charge multiplier for that variable

### Fitting the GLM

```{r glm}
glm_model <- glm(charges ~ age + bmi + smoker + sex + 
                             children + region,
                 data = insurance,
                 family = Gamma(link = "log"))

summary(glm_model)
```

### Interpreting the coefficients

```{r glmcoef}
exp(coef(glm_model))
```

The exponentiated coefficients are charge multipliers:

- `exp(intercept)` = baseline charge (female, non-smoker, 
  age 0, BMI 0, no children, northeast region, not 
  meaningful on its own but anchors the model)
- `exp(smokeryes)` = how many times more a smoker pays 
  vs a non-smoker, holding everything else constant
- `exp(age)` = multiplier per additional year of age
- `exp(bmi)` = multiplier per additional BMI unit

### Observations

The smoker multiplier is 4.48, higher than what was observed in the data. Hence, the GLM is controlling for confounding between smoking and other variables (age, BMI) that the raw comparison, or the decision tree did not account for, because they are not equipped to do it. Once those are controlled for, smoking is even more expensive than the raw ratio suggested.

Age and BMI both have continuous significant effects (p < 0.001): 2.9% per year of age, 1.4% per BMI unit. The tree's age threshold of 43 was thus a computational convenience.

Sex is not significant (p = 0.129) once other variables are controlled. However, it is known that men and women have significantly different mortalities. The absence 
of sex here is dataset-specific as this is not mortality data.

The GLM finds both children and region significant 
(p < 0.001, 8.8% per dependent, southeast and 
southwest significantly cheaper than northeast 
(~13-14%, p < 0.01)) but the tree ignored 
it entirely. This is not a contradiction.

The tree optimises variance reduction at each split. Both children and region have too few unique values to provide any discriminatory ability. The tree found the single best question that separates high vs low cost at a particular node. The GLM finds the marginal effect of each variable controlling for all others.

For my subsequent planned uses, the GLM is more usable because it uses 
all variables simultaneously and produces defensible coefficients. 

#### Predicted charges for our personas

Once we have the GLM coefficients we can compute a 
predicted charge for each persona directly from the 
model.

```{r glmpred}
personas_vbt$predicted_charges <- predict(glm_model, 
  newdata = data.frame(
    age      = personas_vbt$age,
    bmi      = personas_vbt$bmi,
    smoker   = personas_vbt$smoker,
    sex      = personas_vbt$sex,
    children = 0,
    region   = "northeast"
  ), 
  type = "response")


personas_vbt$a_e_ratio = round(personas_vbt$health_charges/personas_vbt$predicted_charges, 3)

personas_vbt
```


This predicted charge becomes the basis for 
their life insurance premium in the next part planned: not the raw 
observed charge, but the model's estimate of what 
someone with their risk profile should cost. Three of the four individuals cost less than expected, but what is interesting is that both the women have a significantly low A to E ratio. 

Elena is obese, a factor that the DT did not consider, and put her into Preferred Elite (notably, the BMI factor was decisive only in the smoking branch in the DT). The GLM looks at the obesity status and predicts a higher premium. 

Caroline is a non-obese smoker. The GLM applies the full smoking multiplier (4.48x) which may overstate costs for a lighter smoker with otherwise normal health indicators.

For both Damon and Niklaus, the predictions are off about 5% over and under actual charges respectively.


## But the predictions were not accurate...

The GLM's predictions are based on average effects across all 1,338 people. Individual variation around those predictions is expected and normal. The A/E ratios here (0.61 to 1.05) show the range of individual variation for typical representatives of each risk class. 

In a real experience study, we would aggregate these individual A/E ratios across thousands of people. The individual noise averages out. What remains is the signal showing whether the risk class was priced correctly on average.

