---
title: "MortaliTeach - build_log"
author: "Debasrija"
date: "2026-05-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)

setwd("/home/ubunt/Downloads/mortaliteach/")
insurance = read.csv("data/insurance.csv")
```

## Step 1 - Data and personas

## Why this project exists

Life and Health insurance pricing is built on this question: given 
everything we know about a person, how likely are they to cost us 
money? The Experience Study is the feedback loop that tells an 
insurer whether their answer to that question was right.

This project builds a tool that teaches that feedback loop from 
first principles, from a single scatter plot to a renegotiated 
reinsurance treaty, using real data and production-grade R engineering.

## What this project teaches

A person who completes this app will understand how insurance 
premiums are derived from observed data, how a life insurance 
pool works, what an Experience Study measures, why count A/E and 
amount A/E can tell different stories, and why reinsurers exist.

## Why this dataset specifically

### What the data is

1,338 real US health insurance records. Age, sex, BMI, smoking 
status, number of children, region, and actual charges billed.

Used to answer: what drives premium differences?

### What I found

- 1,338 rows, 7 columns, no missing values
- Charges: min $1,122, median $9,382, mean $13,270, max $63,770
- Right-skewed, so a small number of very expensive cases pull the mean up
- Age range 18–64
- Approximately equal male-female split, 662 female and 676 male

### What causes the difference in premiums?

One thing that affects premiums significantly is age. Premiums generally go up with age as they are considered higher risk in terms of both life and health.

```{r scatter}
ggplot(data = insurance) + geom_density(aes(x = charges)) + theme_minimal()
ggplot(data = insurance) + geom_point(aes(x = age, y = charges)) + theme_minimal()
```

Interestingly - while the general trend of premiums going up holds, there are 3 bands, which means there is/are some other factor/factors behind the trend, not just age.

#### Smoker split

Based on background reading, I also knew that insurance premiums vary widely depending on whether the customer is a smoker or not, as it affects the mortality measures. So, I decided to start with that.

- There are 274 (20.48%) smokers and 1064 non-smokers. No missing values.
- Mean charges among smokers is $32050.23 and non-smokers is $8434.27, around 3.8x.

Hence, 1,338 real insurance premiums show that smokers pay nearly 3.8 times the premium as compared to a non-smoker.


#### Split by smoking status

```{r scattersmoke}

ggplot(data = insurance) + geom_point(aes(x = age, y = charges, colour = smoker)) +
  scale_color_brewer(palette = "Set1") + theme_minimal()
```
The gap seems partially resolved. The top and bottom bands are composed entirely of smokers and non-smokers, but the middle band is a mixture. There may be other factors which have not been accounted for yet.

#### Split by smoking and obesity status

Another possible split might be BMI (obese/non-obese). Obesity and other comorbidities push premiums high up significantly as well.

```{r scattersmokebmi}

insurance$obese[insurance$bmi >= 30] = "yes"
insurance$obese[insurance$bmi < 30] = "no"

ggplot(data = insurance) + geom_point(aes(x = age, y = charges, colour = smoker, shape = obese), size = 2) +
  scale_color_brewer(palette = "Set1") + theme_minimal()
```

Adding obesity (BMI >= 30) as shape resolves the remaining 
structure. Four groups are now visible:

- Smoker + obese: highest charges, wide spread
- Smoker + non-obese: high charges, below obese smokers  
- Non-smoker + obese: middle band
- Non-smoker + non-obese: bottom band, tight and predictable

Smoking is the dominant signal. Obesity is secondary but 
real-it creates the middle band that was unexplained.

### The four groups

Breaking down by both smoking status and obesity (BMI >= 30):

| Group                    | Mean charges | 
|--------------------------|-------------|
| Smoker + obese           | $41,558     |
| Smoker + non-obese       | $21,363     |
| Non-smoker + obese       | $8,843      |
| Non-smoker + non-obese   | $7,977      |

Two things stand out immediately:

**Smoking dominates obesity.** A smoker with normal BMI 
($21,363) costs more than twice a non-smoker who is obese 
($8,843). The smoking effect overwhelms the BMI effect.

**Obesity matters much more when combined with smoking.** 
The obese/non-obese ratio among smokers is 1.95x ($41,558 
vs $21,363). Among non-smokers it is only 1.11x ($8,843 vs 
$7,977). Obesity and smoking interact - they are not 
simply additive.

These four groups map to risk classes as follows:

- Non-smoker + non-obese  - Preferred Elite
- Non-smoker + obese      - Preferred
- Smoker + non-obese      - Substandard
- Smoker + obese          - Substandard (higher loading)

Smoking alone is disqualifying for Standard or better in real underwriting. Both smoker groups sit in Substandard. The obesity factor makes the combined group more expensive but does not change the classification tier: it changes the premium multiplier within that tier.

I can build a decision tree with this information, grouping subjects into further groups depending on whether this exploration missed any other factors.

So, this mapping may be potentially revised 
when the decision tree is fitted. The tree may suggest 
different natural groupings.

The first iteration of the decision tree is here - 

```{r decisiontree}
library(rpart)
library(rpart.plot)

insurance$smoker <- as.factor(insurance$smoker)
insurance$sex    <- as.factor(insurance$sex)
insurance$region <- as.factor(insurance$region)

tree <- rpart(charges ~ age + bmi + smoker + sex + children + region,
              data = insurance,
              method = "anova")

importance <- as.data.frame(tree$variable.importance)
importance$variable <- rownames(importance)
colnames(importance)[1] <- "importance"

ggplot(importance, aes(x = reorder(variable, importance), 
                       y = importance)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Variable importance in predicting charges",
       x = NULL, y = "Importance") +
  theme_minimal()
```

This shows that smoking status, BMI and age are the three main factors behind the split in charges. I did not add obese as a dichotomous variable since I wanted to see where the split naturally happens given the dataset. (Note for self: charges is a continuous outcome variable, so the method used was ANOVA.)

```{r dt}
printcp(tree)
rpart.plot(tree, type = 4, extra = 101, roundint = FALSE)
```

Within non-smokers, age is more discriminating than obesity at this level of the tree. The younger non-smokers average $5,399, older ones $12,000. Within smokers, the dominant factor becomes obesity, with the premiums being higher than non-smokers all around, non-obese individuals averaging $21,000 and obese averaging $42,000.

Another interesting thing is that the BMI categorisation for smokers does happen at the exact thresholds where obesity is defined. This serves as an extra sanity check as to whether the thresholds used earlier were correct given the dataset.

#### Why sex, region, and children were not used

Sex, region, and children are not predictive of charges once smoking, age, and BMI are known. This is a finding, not a data problem.

In real insurance underwriting, sex does affect pricing (women generally have lower mortality than men). The fact that it does not appear here is likely because this is a
health insurance charges dataset, not a life insurance mortality dataset. 


The next thing to do now is to pick one representative persona for each of these groups. I decided to pick the individuals closest to the mean.


```{r persona}
insurance$leaf <- tree$where

leaf_means <- aggregate(charges ~ leaf, data = insurance,
                        FUN = mean)

personas <- do.call(rbind, lapply(unique(insurance$leaf), function(l) {
  group <- insurance[insurance$leaf == l, ]
  leaf_mean <- leaf_means$charges[leaf_means$leaf == l]
  closest <- which.min(abs(group$charges - leaf_mean))
  group[closest, ]
}))

print(personas)
```

These individuals are real ones picked from the dataset, and reduces the amount of ambiguity in deciding what characteristics should be present in these personas when designed. 

Since I am planning to make an app, I will assign names to these individuals.

Now we give them names: (_wild Vampire Diaries reference appears_...)

- Row 524: **Elena** - 38, female, non-smoker, obese. 
  Preferred Elite despite high BMI because she does not smoke 
  and is under 43. Pays $5,398.

- Row 842: **Damon** - 59, male, non-smoker, normal BMI. 
  Preferred as older age moves him out of Preferred Elite. 
  Pays $12,324.

- Row 161: **Caroline** - 42, female, smoker, normal BMI. 
  Substandard. Smoking alone puts her here regardless of 
  normal BMI. Pays $21,349.

- Row 1091: **Niklaus** - 47, male, smoker, obese. 
  Substandard high loading. Both risk factors present. 
  Pays $41,676.

These four people will appear in every section of the app. 
Their numbers are real. Their names are not. Overall, the message here is that the premium order is driven almost entirely by smoking. This is not a biological truth, but is true for this dataset.

Making a dataframe for these 4 individuals.

```{r personadt}
# VBT 2015 central death rates for the four personas
# Source: SOA Valuation Basic Table 2015
# mx = central death rate (deaths per person per year)

personas_vbt <- data.frame(
  name   = c("Elena", "Damon", "Caroline", "Niklaus"),
  age    = c(38, 59, 42, 47),
  sex    = c("female", "male", "female", "male"),
  smoker = c("no", "no", "yes", "yes"),
  bmi    = c(37.73, 24.70, 26.60, 36.19),
  health_charges = c(5397.617, 12323.936, 21348.706, 41676.081),
  risk_class = c("Preferred Elite", "Preferred", 
                 "Substandard", "Substandard High"),
  stringsAsFactors = FALSE
)
```

ANB is used as the best approximation of age since it generally gives a more precise result. 

From the 2015 VBT lookup table - we can get qx, and assuming uniform force of mortality, mx can be calculated. 

```{r qx}
personas_vbt$qx <- c(
  0.00012, # Elena: F38NS
  0.00378, # Damon: M59NS  
  0.00045, # Caroline: F42S
  0.00323 # Niklaus: M47S
)

personas_vbt$mx = -log(1 - personas_vbt$qx)

print(personas_vbt)
```


Mortality vs charges ratio at similar ages should be consistent, in the same direction at least.
```{r check}
# mortality ratio: smoker vs non-smoker at similar ages
cat("Caroline/Elena mortality ratio:", 
    round(personas_vbt$mx[3] / personas_vbt$mx[1], 2), "\n")

cat("Health charges ratio (smoker/non-smoker):", 
    round(21348 / 5397, 2), "\n")
```

Smoker/non-smoker mortality ratio (3.75x) is consistent 
with the health charges ratio from insurance data (3.96x). 
This confirms the bridge between the two datasets - the 
same variables that predict health costs predict mortality. 

### Why a GLM alongside the decision tree

The decision tree showed us the risk classes visually. It is readable, intuitive, and derives groupings from data without any prior assumptions. But it has some weaknesses that make it unsuitable as a standalone model:

1. Small changes in the data can change the splits entirely.
2. The decision tree cannot adjust for factors like age and BMI between the smokers and non-smokers.
3. Age splits at exactly 43, but the true relationship between age and charges is continuous so it is not explainable fully.
4. No uncertainty estimates like standard errors, or confidence intervals.

The GLM addresses all three. It models the continuous 
relationship between each variable and charges, produces 
standard errors on every coefficient, and is stable across 
small data changes.

### Why Gamma with log link

Charges are strictly positive and right-skewed. A standard 
linear regression (Gaussian GLM) would allow negative 
predicted values, which is impossible for insurance charges. The 
Gamma distribution respects the constraint that charges > 0 
and matches the shape of the observed distribution.

The log link models multiplicative effects. A smoker does 
not pay a fixed dollar amount more than a non-smoker, they pay a multiple of what a non-smoker pays. The log link captures this correctly. It also means coefficients 
are directly interpretable as multipliers when exponentiated:

exp(coefficient) = charge multiplier for that variable

### Fitting the GLM

```{r glm}
glm_model <- glm(charges ~ age + bmi + smoker + sex + 
                             children + region,
                 data = insurance,
                 family = Gamma(link = "log"))

summary(glm_model)
```

### Interpreting the coefficients

```{r glmcoef}
exp(coef(glm_model))
```

The exponentiated coefficients are charge multipliers:

- `exp(intercept)` = baseline charge (female, non-smoker, 
  age 0, BMI 0, no children, northeast region, not 
  meaningful on its own but anchors the model)
- `exp(smokeryes)` = how many times more a smoker pays 
  vs a non-smoker, holding everything else constant
- `exp(age)` = multiplier per additional year of age
- `exp(bmi)` = multiplier per additional BMI unit

### Observations

The smoker multiplier is 4.48, higher than what was observed in the data. Hence, the GLM is controlling for confounding between smoking and other variables (age, BMI) that the raw comparison, or the decision tree did not account for, because they are not equipped to do it. Once those are controlled for, smoking is even more expensive than the raw ratio suggested.

Age and BMI both have continuous significant effects (p < 0.001): 2.9% per year of age, 1.4% per BMI unit. The tree's age threshold of 43 was thus a computational convenience.

Sex is not significant (p = 0.129) once other variables are controlled. However, it is known that men and women have significantly different mortalities. The absence 
of sex here is dataset-specific as this is not mortality data.

The GLM finds both children and region significant 
(p < 0.001, 8.8% per dependent, southeast and 
southwest significantly cheaper than northeast 
(~13-14%, p < 0.01)) but the tree ignored 
it entirely. This is not a contradiction.

The tree optimises variance reduction at each split. Both children and region have too few unique values to provide any discriminatory ability. The tree found the single best question that separates high vs low cost at a particular node. The GLM finds the marginal effect of each variable controlling for all others.

For my subsequent planned uses, the GLM is more usable because it uses 
all variables simultaneously and produces defensible coefficients. 

#### Predicted charges for our personas

Once we have the GLM coefficients we can compute a 
predicted charge for each persona directly from the 
model.

```{r glmpred}
personas_vbt$predicted_charges <- predict(glm_model, 
  newdata = data.frame(
    age      = personas_vbt$age,
    bmi      = personas_vbt$bmi,
    smoker   = personas_vbt$smoker,
    sex      = personas_vbt$sex,
    children = 0,
    region   = "northeast"
  ), 
  type = "response")


personas_vbt$a_e_ratio = round(personas_vbt$health_charges/personas_vbt$predicted_charges, 3)

personas_vbt
```


This predicted charge becomes the basis for 
their life insurance premium in the next part planned: not the raw 
observed charge, but the model's estimate of what 
someone with their risk profile should cost. Three of the four individuals cost less than expected, but what is interesting is that both the women have a significantly low A to E ratio. 

Elena is obese, a factor that the DT did not consider, and put her into Preferred Elite (notably, the BMI factor was decisive only in the smoking branch in the DT). The GLM looks at the obesity status and predicts a higher premium. 

Caroline is a non-obese smoker. The GLM applies the full smoking multiplier (4.48x) which may overstate costs for a lighter smoker with otherwise normal health indicators.

For both Damon and Niklaus, the predictions are off about 5% over and under actual charges respectively.


## But the predictions were not accurate...

The GLM's predictions are based on average effects across all 1,338 people. Individual variation around those predictions is expected and normal. The A/E ratios here (0.61 to 1.05) show the range of individual variation for typical representatives of each risk class. 

In a real experience study, we would aggregate these individual A/E ratios across thousands of people. The individual noise averages out. What remains is the signal showing whether the risk class was priced correctly on average.

The A/E ratios we got were per person, but for a proper actuarial study, one would scale up to many more individuals, follow them over a period of time and see how they fare over this period. To scale this up, I went for a modular approach that separates the analysis functions to make them testable in silo. This is the analytics engine that will come in this iteration.

## Part 2 - Analytics engine

### Scope

Functions to build:
- `validate_cohort()` - hard failures on bad input data
- `.load_nhanes()` - download and clean NHANES data
- `load_cohort()` - public wrapper around `.load_nhanes()` (validates the cohort)
- `build_exposure()` - expand cohort to person-year rows, needed format for the study
- `attach_basis()` - join VBT mortality rates to exposure
- `run_study()` - compute A/E

For this, I used the NHANES data.

### Dataset decision-why continuous NHANES 1999-2004

Used NHANES 1999-2004 (three waves combined).

```
Wave 1: ~5,000 adults, examined 1999-2000, up to 20yr follow-up
Wave 2: ~5,400 adults, examined 2001-2002, up to 18yr follow-up  
Wave 3: ~5,000 adults, examined 2003-2004, up to 16yr follow-up
```

**Why continuous NHANES 1999-2004:**
- nhanesA handles data retrieval cleanly - age, sex, smoking all 
  available through one function call
- Mortality linkage files parse correctly with the CDC fixed-width format recommended
- Three waves give ~15,000 eligible adults with 16-20 years of follow-up-enough deaths for credible A/E ratios

**Tradeoffs:**
- Basis mismatch: observed cohort 1999-2004 vs VBT 2015 expected rates-expected to show A/E near 1.0 or slight improvement due to mortality trends. However, this can be used to show if the basis is adequate.
- This mismatch is itself a teaching moment-the 
  experience study reveals whether the basis is adequate

### Why three waves specifically

Adding waves beyond 2003-2004 introduces changing smoking 
prevalence and mortality trends that complicate the teaching 
narrative without adding proportionate value.

### Data structure-two files per wave

The CDC splits participant information across two sources:

**Survey file** (nhanesA package):
- DEMO has SEQN, RIDAGEYR (age in years), RIAGENDR (sex)
- SMQ has SEQN, SMQ020 (ever smoked 100 cigarettes)

**Mortality linkage file** (CDC fixed-width - three files 1999-00, 2001-02, 2003-04):
- SEQN, eligstat, mortstat, permth_exm (follow-up months)

Joined on SEQN. Neither file is complete alone.

### SMQ020 missingness investigation

After joining DEMO to SMQ for 1999-2000 wave:
- 4,517 non-adults (age < 18): no SMQ020 - expected, smoking questionnaire not administered to children
- 4,877 adults: have SMQ020 
- 571 adults: missing SMQ020

Of the 571 missing adults:
- 37 interviewed only, never examined, and so will not appear 
  in mortality linkage file, drop out on join
- 534 examined but no smoking status: almost entirely 
  age 18-19, likely questionnaire age threshold issue

**Decision: exclude all participants with missing SMQ020.**
No imputation as the smoker-distinct analysis requires known 
smoking status for every participant.

**Additional filter:** keep only examined participants
(RIDSTATR == "Both Interviewed and MEC examined") to 
ensure presence in mortality linkage file.

### validate_cohort() design and tests

Hard failures on six conditions. Returns df invisibly 
if all pass, stops with precise error message if any fail.

### Package dependencies-analytics.R

data.table requires library() for the [ operator-
cannot be fully namespaced. One library() call at 
top of file. All other packages use explicit :: 
namespacing throughout.

Known technical debt: library(data.table) will be 
replaced with @import data.table in roxygen comments 
when converting to golem package in Iteration 4.

Dependencies:
- data.table: exposure frame construction, keyed joins
- readr: fixed-width mortality file parsing  
- nhanesA: NHANES survey data download
- dplyr: cohort cleaning
- tidyr: unnest in profiling version only

```{r validate_cohort_tests}
source("R/analytics.R")

good <- data.frame(
  seqn            = 1L,
  age_at_exam     = 52L,
  sex             = "Male",
  smoker          = "non_smoker",
  vital_status    = 1L,
  follow_up_years = 8.7,
  stringsAsFactors = FALSE
)

validate_cohort(good)                    # passes silently

bad_age <- good; bad_age$age_at_exam <- 150L
tryCatch(validate_cohort(bad_age), 
         error = function(e) message(e$message))

bad_vs <- good; bad_vs$vital_status <- 9L
tryCatch(validate_cohort(bad_vs),  
         error = function(e) message(e$message))

bad_sex <- good; bad_sex$sex <- "M"
tryCatch(validate_cohort(bad_sex), 
         error = function(e) message(e$message))

bad_fu <- good; bad_fu$follow_up_years <- -1
tryCatch(validate_cohort(bad_fu),  
         error = function(e) message(e$message))

bad_na <- good; bad_na$age_at_exam <- NA
tryCatch(validate_cohort(bad_na),  
         error = function(e) message(e$message))

bad_smoker <- good; bad_smoker$smoker <- "yes"
tryCatch(validate_cohort(bad_smoker),
         error = function(e) message(e$message))
```

All six checks pass. Function complete.

### Why each check exists

**Missing columns:** downstream functions use column names 
directly-a missing column produces a cryptic error six 
functions deep. Better to fail at the door with a clear 
message.

**NAs in required columns:** NA in age_at_exam would 
produce NA rows in the exposure frame. NA in vital_status 
would corrupt the death count. NA in follow_up_years would 
produce NA exposure. All silent corruptions.

**Age 18-85:** NHANES exam ages cap at 85 in practice.
Maximum attained age during follow-up is 104.5 years -
the VBT 2015 covers this (goes to age 120). The 85 limit
applies at entry only, not during follow-up.

**follow_up_years > 0:** Zero or negative follow-up 
produces zero or negative exposure rows. The conservation 
law test would still pass but the person contributes 
nothing to the study-they should not be in the cohort.

**vital_status in {0,1}:** Any other value means the 
linkage failed or the data was corrupted. 2, 7, 9 are 
common NHANES codes for refused/unknown-these must 
never be silently treated as alive or dead.

**sex in {Male, Female}:** attach_basis() joins on sex 
to get the correct VBT rate. Any other value produces 
NA ref_rate for that participant-the silent NA 
corruption that must never reach run_study().

### Cohort summary-load_cohort() output

```
14,170 participants-three waves combined
Female: 7,458 (52.6%)
Male:   6,712 (47.4%)
Non-smoker: 7,295 (51.5%)
Smoker:     6,875 (48.5%)
Deaths: 4,087 (28.8%)
Age at exam: median 48, range 20-85
Follow-up: median 16.8 years, max 20.75 years
```

We observed that out of the eligible ones, 48.5% are smokers. This is high compared to the ~20% in insurance.csv. This reflects two things, that NHANES 1999-2004 had higher smoking prevalence than current rates, and "ever smoked 100 cigarettes" includes former smokers who may have quit. Former smokers may have lower mortality than current smokers. The VBT smoker rates assume current smokers. This will likely push smoker A/E
below 1.0.This will affect our A/E ratios and is worth documenting.


### profvis results-build_exposure_loop

Flame graph shows data.frame() as the dominant cost 
(orange bar, second widest after the outer function).

Root cause confirmed: constructing one data.frame per 
participant inside the loop triggers 14,170 separate 
memory allocations. Each allocation increases GC 
pressure. The arithmetic (seq, rep, floor) is 
essentially free by comparison.

This is why dplyr rowwise() did not fix the problem-
it also allocates per-row intermediates, just with 
cleaner syntax hiding the cost.

data.table fixes the root cause: rep() replicates the 
entire table in one allocation, then := mutates in-place 
with no copying. Zero per-row allocations.

Timings on 14,170 NHANES participants:

- Loop:        4.76s
- dplyr:       2.10s  
- data.table:  0.027s
- Speedup:     176x (loop), 78x (dplyr)


### build_exposure()-conservation law results

Verified on all three implementations against 14,170 
NHANES participants:

**Conservation law 1-exposure totals:**
sum(exposure$exposure_yrs) == sum(cohort$follow_up_years)
Loop, dplyr, and data.table all equal. Pass.

**Conservation law 2-death totals:**
sum(exposure$died) == sum(cohort$vital_status) == 4,087
All three equal. Pass.

All three implementations produce identical results.
Performance difference is purely allocations, not 
correctness. So now we know that the code is not logically incorrect - these approaches all do the same thing, in different ways, at different efficiencies. 
The data.table approach is the correct one here. It is great in terms of parallelization, as we saw, it sped things up massively. The speedup 
comes from eliminating per-row memory allocation - 
not from parallelisation. The loop creates 14,170 
separate data frame objects triggering repeated GC. 
data.table creates one object and mutates in-place 
with := - zero copies, zero per-row allocation.

data.table does use multiple threads for grouped 
aggregations, which will benefit run_study(). But 
the build_exposure() speedup is purely allocation 
efficiency.

### attach_basis() - first A/E results

Overall A/E: 1.234 - actual mortality 23% above VBT 2015 
expected.

By smoker status:
- Non-smoker A/E: 1.329
- Smoker A/E:     1.175

By sex:
- Female A/E: 1.226
- Male A/E:   1.241

**Interpretation:** Overmortality is concentrated in 
non-smokers and is symmetric across sexes. This is 
consistent with a select vs ultimate population effect:

VBT 2015 non-smoker rates are calibrated on insured 
lives who passed underwriting - a healthier-than-average 
population. NHANES non-smokers are general population 
including people with poor health outcomes who would 
not qualify for insurance.

This is expected and is itself a teaching moment: 
experience studies reveal basis mismatches. The VBT 
is not wrong - it is calibrated for a different 
population than NHANES. A real insurer would expect 
A/E near 1.0 on their own book of insured lives.

The 1.234 overall A/E will drive the repricing 
discussion in Act 3 - the treaty needs renegotiation.


### run_study() - results

Four groups, all reliable (A >= 5):

| Group | A | E | A/E | 95% CI |
|---|---|---|---|---|
| Female non-smoker | 1043 | 782 | 1.334 | 1.105–1.591 |
| Female smoker | 834 | 749 | 1.114 | 0.902–1.356 |
| Male non-smoker | 654 | 495 | 1.321 | 1.039–1.647 |
| Male smoker | 1556 | 1286 | 1.210 | 1.038–1.399 |

Three of four CIs entirely above 1.0 - overmortality 
is statistically significant, not noise.

Female smoker CI crosses 1.0 - cannot reject null 
hypothesis of A/E = 1.0 for this group at 95% confidence.

Non-smoker A/E consistently higher than smoker A/E 
in both sexes - confirms select population effect. 
VBT non-smoker rates calibrated on healthier insured 
lives; NHANES non-smokers are general population.

**Teaching narrative for Act 3:**
The treaty was priced using VBT 2015 rates. Actual 
mortality was 23% above expected overall. The Byar 
CIs confirm this is a real signal. At treaty renewal, 
the reinsurer will argue for higher multipliers - 
especially for non-smoker groups where the basis 
mismatch is largest.


## Iteration 2 - Financial layer

### Scope

Functions to build:
- `assign_risk_class()` - underwriting rules → risk class
- `assign_financials()` - face amount, premium, NAR
- `reinsurance_split()` - retained vs ceded by policy
- `compute_repricing()` - new multipliers from A/E results

### Adding BMI to the cohort

The four-class risk system from Iteration 0 requires BMI
to split smokers into Substandard and Substandard High.
BMI is not in the mortality linkage file - it comes from
the NHANES body measurements component (BMX table).

Added to `.load_nhanes()`:
- Wave 1: `BMX` → `BMXBMI`
- Wave 2: `BMX_B` → `BMXBMI`
- Wave 3: `BMX_C` → `BMXBMI`

**Missingness check - wave 1 after existing filters:**
- 4,435 adults with known smoking status
- 68 missing BMI (1.5%)
- Decision: exclude. No imputation - 1.5% loss is
  acceptable and imputing BMI for risk classification
  would introduce assumption-dependent risk classes.

**Updated cohort after adding BMI:**
- 13,675 participants (down from 14,170)
- 495 dropped due to missing BMI across three waves
- BMI range: 12.04 – 66.44, median 27.37
- No NAs remaining

Added BMI validation to `validate_cohort()`:
- Column must be present
- Values must be between 10 and 80
- No NAs permitted

### assign_risk_class() - design

Rules derived from the Iteration 0 decision tree.
The tree found: smoking first, then age for non-smokers,
then BMI for smokers.

```
Non-smoker, age < 43   → Preferred Elite   (1.00x)
Non-smoker, age >= 43  → Preferred         (1.25x)
Smoker, BMI < 30       → Substandard       (2.33x)
Smoker, BMI >= 30      → Substandard High  (3.25x)
```

Multipliers are relative to Preferred Elite baseline.
Derived from the mx ratios between VBT risk groups,
rounded to practical values.

**Why these multipliers:**
- Preferred to Preferred Elite: age split within
  non-smokers. Older non-smokers cost 1.25x more
  on average based on the leaf means from the tree
  ($12,300 vs $5,399 = 2.28x health charges, but
  mortality ratio is smaller - 1.25x reflects the
  VBT age gradient not the health charge gradient).
- Substandard to Preferred Elite: smoker non-obese
  vs young non-smoker. mx ratio from VBT confirms
  roughly 2.3x mortality loading for smokers.
- Substandard High to Preferred Elite: smoker obese
  vs young non-smoker. Additional BMI loading on top
  of smoking brings this to 3.25x.

```{r assign_risk_class}
cohort <- assign_risk_class(cohort)
table(cohort$risk_class)
```

### What the risk class distribution shows

[fill in after running - paste table(cohort$risk_class) output]

### Connection to personas

The four NHANES risk classes map directly to the four
personas from Iteration 0:

| Persona | Age | Smoker | BMI | Risk class |
|---|---|---|---|---|
| Elena | 38 | No | 37.73 | Preferred Elite |
| Damon | 59 | No | 24.70 | Preferred |
| Caroline | 42 | Yes | 26.60 | Substandard |
| Niklaus | 47 | Yes | 36.19 | Substandard High |

The personas were selected to be typical representatives
of each leaf in the decision tree. The NHANES cohort
applies the same rules to 13,675 real participants.

### Risk class distribution

```
Preferred Elite:  3,346 (24.5%)
Preferred:        3,700 (27.1%)
Substandard:      4,574 (33.4%)
Substandard High: 2,055 (15.0%)
```

Substandard is the largest group (33.4%) reflecting 
the high smoking prevalence in the 1999-2004 cohort. 
Preferred Elite is the smallest non-Substandard group 
- only non-smokers under 43 qualify.

Spot checks confirmed:
- Non-smoker age 38 - Preferred Elite 
- Smoker age 47, BMI >= 30 - Substandard High 


### assign_financials() - design

Premium formula:

```
annual_premium = $1,200 (fixed budget for all participants)
face_amount    = 1200 / (mx × multiplier × 1.15)
NAR            = face_amount
```

The fixed budget approach makes the Preferred Elite 
insight concrete - the same $1,200 buys vastly more 
cover for low-risk lives because their mx is low.

NAR is simplified to face amount. A real insurer would 
subtract the policy reserve from face amount to get 
true net amount at risk. We use face amount directly 
as a teaching simplification - documented here so it 
is not mistaken for a production implementation.

Expense loading of 1.15 covers admin costs and profit 
margin (15%).

### assign_financials() - results

```{r financials}
aggregate(face_amount ~ risk_class,
          data = cohort,
          FUN  = function(x) round(mean(x)))
```

| Risk class | Mean face amount |
|---|---|
| Preferred Elite | $8,303,017 |
| Preferred | $295,125 |
| Substandard | $569,447 |
| Substandard High | $398,497 |

Preferred Elite buys 28x more cover than Preferred 
for the same $1,200 annual budget. This is the central 
teaching insight of Act 2 - the same premium buys 
vastly different cover depending on mortality risk.

### Face amount inversion - Substandard vs Substandard High

Observed: mean face amount for Substandard ($569,447) 
is higher than Substandard High ($398,497) despite 
Substandard High having a higher multiplier (3.25x vs 
2.33x).

Investigation:
- Mean age nearly identical: 51.4 vs 51.0 years
- Mean mx: Substandard 0.01578, Substandard High 0.01144

Root cause: the VBT 2015 smoker rates do not distinguish 
by BMI. A smoker with BMI 25 and a smoker with BMI 35 
get the same mx at the same age and sex. The BMI split 
creates two groups from the same rate table - the mx 
difference between them reflects only small age 
distribution differences within each class.

The multiplier difference (2.33x vs 3.25x) between 
Substandard and Substandard High is derived from 
health insurance charges data (Iteration 0), not from 
VBT mortality data. Health charges show obese smokers 
cost materially more than non-obese smokers ($41,558 
vs $21,363). The VBT does not reflect this distinction.

**Decision: keep four classes and document the mismatch.**

This surfaces a real tension in insurance pricing:
- Health charges data: BMI materially affects cost 
  for smokers
- VBT mortality data: BMI does not affect mortality 
  for smokers in the published table

Underwriting in practice uses multiple data sources. 
The four-class system reflects health cost evidence 
even where mortality evidence is absent. This is a 
documented assumption, not a data error.

**Teaching moment for Act 3:** The experience study 
will show whether Substandard High actually has worse 
mortality than Substandard in the NHANES cohort - 
real data test of the assumption.

### reinsurance_split() - design

The insurer retains the first $500,000 of each policy. 
Anything above is ceded to the reinsurer.

```
retained_nar    = min(nar, 500,000)
ceded_nar       = max(nar - 500,000, 0)
cession_rate    = ceded_nar / nar
retained_premium = annual_premium × (1 - cession_rate)
ceded_premium   = annual_premium × cession_rate
treaty_premium  = ceded_nar × mx × multiplier × 1.10
```

Treaty loading of 1.10 - the reinsurer charges 10% 
loading on the ceded risk. Lower than the direct 
insurer's 15% loading because the reinsurer has lower 
expense base and diversifies across many cedants.

The retention limit of $500,000 is typical for a 
small-medium insurer. Preferred Elite policies 
(mean face $8.3M) will cede most of their NAR. 
Preferred policies (mean face $295K) will retain 
everything below $500K - most are fully retained.

**Why reinsurance exists - the teaching narrative:**

The Preferred Elite policies buy enormous face amounts 
($8-20M range) for $1,200/year because their mortality 
is very low. But if one of those lives dies, the claim 
is $8-20M. A single death in that group could bankrupt 
a small insurer.

The reinsurer absorbs the tail risk above $500K per 
policy. The insurer pays a treaty premium for this 
protection. The experience study in Act 3 will show 
whether the treaty was priced correctly - did the 
reinsurer receive enough premium to cover their 
ceded claims?

### reinsurance_split() — results

```
Total NAR:    $32.3 billion
Retained NAR:  $4.0 billion (12.4%)
Ceded NAR:    $28.3 billion (87.6%)
Mean cession:  34.1% per policy
```

By risk class (mean per policy):

| Risk class | Retained | Ceded | Total NAR |
|---|---|---|---|
| Preferred Elite | $500K | $7,803K | $8,303K |
| Preferred | $239K | $56K | $295K |
| Substandard | $223K | $347K | $569K |
| Substandard High | $200K | $198K | $398K |

Preferred Elite is almost entirely reinsured — mean 
ceded NAR of $7.8M per policy vs retention of $500K. 
The reinsurer holds 94% of Preferred Elite risk.

Preferred is mostly retained — mean face $295K is 
below the $500K retention limit for most policies.

This confirms the teaching narrative: the insurer 
cannot hold Preferred Elite risk alone. A single 
Preferred Elite death produces a claim of $8M+ 
against $1,200 in annual premium. The reinsurer 
exists to absorb this tail risk.

### run_study() by risk class — results

Overall A/E: 1.192 (A = 3,801, E = 3,189)

| Risk class | A | E | A/E | 95% CI | Reliable |
|---|---|---|---|---|---|
| Preferred Elite | 100 | 35 | 2.834 | 1.467–4.824 | Yes |
| Preferred | 1,470 | 1,187 | 1.238 | 1.057–1.437 | Yes |
| Substandard | 1,555 | 1,405 | 1.107 | 0.950–1.280 | Yes |
| Substandard High | 676 | 562 | 1.203 | 0.950–1.495 | Yes |

**Headline finding: Preferred Elite A/E = 2.834**

Actual mortality was 183% above expected for the 
lowest-risk group. CI entirely above 1.0 — 
statistically significant.

Explanation: VBT non-smoker rates for young adults 
are calibrated on healthy insured lives. NHANES young 
non-smokers are general population including people 
who would never qualify for insurance. The basis 
mismatch is largest for the group with the lowest 
absolute mortality rate — a small absolute difference 
produces a large A/E ratio when E is tiny.

**Substandard and Substandard High CIs cross 1.0** — 
cannot reject A/E = 1.0 for smoker groups. The VBT 
smoker rates, calibrated on a more general population, 
are adequate for the NHANES smoker cohort.

**Teaching narrative for Act 3:**
The treaty was priced assuming Preferred Elite was 
the safest group. The experience study reveals it 
has the worst A/E ratio. This is the basis mismatch 
made visible — and the reason the reinsurer will 
demand the highest multiplier increase at renewal 
for the Preferred Elite group.

### compute_repricing() — results

Credibility-weighted repricing using full credibility 
threshold of 1,082 deaths (Poisson, 95%/5% precision).

| Risk class | Old mult | A/E | Credibility | New mult | Change |
|---|---|---|---|---|---|
| Preferred Elite | 1.00x | 2.834 | 9.2% | 1.17x | +17% |
| Preferred | 1.25x | 1.238 | 100% | 1.55x | +24% |
| Substandard | 2.33x | 1.107 | 100% | 2.58x | +11% |
| Substandard High | 3.25x | 1.203 | 62.5% | 3.66x | +13% |

**Key insight — credibility dominates repricing:**

Preferred Elite has the worst A/E (2.834) but the 
smallest price increase (+17%) because only 100 deaths 
were observed — 9.2% credibility. The repricing is 
heavily pulled toward 1.0 by the low weight.

Preferred has the largest price increase (+24%) despite 
a lower A/E (1.238) because 1,470 deaths gives full 
credibility. The experience fully drives the new price.

This is not a bug — it is the credibility mechanism 
working correctly. Acting on a 2.834 A/E from 100 
deaths would be overreacting to noise. Waiting for 
more Preferred Elite deaths before a large repricing 
is the actuarially correct response.

**Teaching narrative:**
At treaty renewal, the reinsurer presents this table. 
The insurer expected Preferred Elite to be their 
safest group — it has the worst A/E. But the small 
number of deaths limits how aggressively the reinsurer 
can reprice. Both parties agree to monitor Preferred 
Elite closely and revisit after the next study period.