---
title: "MortaliTeach - build_log"
author: "Debasrija"
date: "2026-05-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)

setwd("/home/ubunt/Downloads/mortaliteach/")
insurance = read.csv("data/insurance.csv")
```

## Step 1 - Data and personas

## Why this project exists

Life and Health insurance pricing is built on this question: given 
everything we know about a person, how likely are they to cost us 
money? The Experience Study is the feedback loop that tells an 
insurer whether their answer to that question was right.

This project builds a tool that teaches that feedback loop from 
first principles, from a single scatter plot to a renegotiated 
reinsurance treaty, using real data and production-grade R engineering.

## What this project teaches

A person who completes this app will understand how insurance 
premiums are derived from observed data, how a life insurance 
pool works, what an Experience Study measures, why count A/E and 
amount A/E can tell different stories, and why reinsurers exist.

## Why this dataset specifically

### What the data is

1,338 real US health insurance records. Age, sex, BMI, smoking 
status, number of children, region, and actual charges billed.

Used to answer: what drives premium differences?

### What I found

- 1,338 rows, 7 columns, no missing values
- Charges: min $1,122, median $9,382, mean $13,270, max $63,770
- Right-skewed, so a small number of very expensive cases pull the mean up
- Age range 18–64
- Approximately equal male-female split, 662 female and 676 male

### What causes the difference in premiums?

One thing that affects premiums significantly is age. Premiums generally go up with age as they are considered higher risk in terms of both life and health.

```{r scatter}

ggplot(data = insurance) + geom_point(aes(x = age, y = charges)) + theme_minimal()
```

Interestingly - while the general trend of premiums going up holds, there are 3 bands, which means there is/are some other factor/factors behind the trend, not just age.

#### Smoker split

Based on background reading, I also knew that insurance premiums vary widely depending on whether the customer is a smoker or not, as it affects the mortality measures. So, I decided to start with that.

- There are 274 (20.48%) smokers and 1064 non-smokers. No missing values.
- Mean charges among smokers is $32050.23 and non-smokers is $8434.27, around 3.8x.

Hence, 1,338 real insurance premiums show that smokers pay nearly 3.8 times the premium as compared to a non-smoker.


#### Split by smoking status

```{r scattersmoke}

ggplot(data = insurance) + geom_point(aes(x = age, y = charges, colour = smoker)) +
  scale_color_brewer(palette = "Set1") + theme_minimal()
```
The gap seems partially resolved. The top and bottom bands are composed entirely of smokers and non-smokers, but the middle band is a mixture. There may be other factors which have not been accounted for yet.

#### Split by smoking and obesity status

Another possible split might be BMI (obese/non-obese). Obesity and other comorbidities push premiums high up significantly as well.

```{r scattersmokebmi}

insurance$obese[insurance$bmi >= 30] = "yes"
insurance$obese[insurance$bmi < 30] = "no"

ggplot(data = insurance) + geom_point(aes(x = age, y = charges, colour = smoker, shape = obese), size = 2) +
  scale_color_brewer(palette = "Set1") + theme_minimal()
```
Adding obesity (BMI >= 30) as shape resolves the remaining 
structure. Four groups are now visible:

- Smoker + obese: highest charges, wide spread
- Smoker + non-obese: high charges, below obese smokers  
- Non-smoker + obese: middle band
- Non-smoker + non-obese: bottom band, tight and predictable

Smoking is the dominant signal. Obesity is secondary but 
real — it creates the middle band that was unexplained.

### The four groups

Breaking down by both smoking status and obesity (BMI >= 30):

| Group                    | Mean charges | 
|--------------------------|-------------|
| Smoker + obese           | $41,558     |
| Smoker + non-obese       | $21,363     |
| Non-smoker + obese       | $8,843      |
| Non-smoker + non-obese   | $7,977      |

Two things stand out immediately:

**Smoking dominates obesity.** A smoker with normal BMI 
($21,363) costs more than twice a non-smoker who is obese 
($8,843). The smoking effect overwhelms the BMI effect.

**Obesity matters much more when combined with smoking.** 
The obese/non-obese ratio among smokers is 1.95x ($41,558 
vs $21,363). Among non-smokers it is only 1.11x ($8,843 vs 
$7,977). Obesity and smoking interact - they are not 
simply additive.

These four groups map to risk classes as follows:

- Non-smoker + non-obese  - Preferred Elite
- Non-smoker + obese      - Preferred
- Smoker + non-obese      - Substandard
- Smoker + obese          - Substandard (higher loading)

Smoking alone is disqualifying for Standard or better in 
real underwriting. Both smoker groups sit in Substandard.
The obesity factor makes the combined group more expensive 
but does not change the classification tier — it changes 
the premium multiplier within that tier.

I can build a decision tree with this information, grouping subjects into further groups depending on whether this exploration missed any other factors.

So, this mapping may be potentially revised 
when the decision tree is fitted. The tree may suggest 
different natural groupings.